The Unicode Cookbook \linebreak for Linguists

### 2018-01-13 book/chapters/ipa_meets_unicode.tex

As long as a transcription is only directed towards phonetically trained eyes,
then all the details of the Unicode-encoding are unimportant. For a linguist
reading an IPA transcription, many of the details that will be discussed in this
chapter might seem like hair-splitting trivialities. However, if IPA
transcriptions are intended to be used across resources (e.g.~searching similar
phenomena across different languages) then it becomes crucial that there are
strict encoding guidelines. Our main goal in this chapter is to present the
encoding issues and propose recommendations for a strict IPA encoding for
situations in which cross-resource consistency is crucial.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
code blocks shown here are the characters proposed for strict IPA encoding, 
as discussed in Section~\ref{ipa-recommendations}.}

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the real retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        bevavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be used, respectively.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is the character to by used for a major group 
           break (by intonation), to be encoded by \textsc{double vertical line} 
           at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different dashed-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable when the goal is interoperability between resources written in
IPA.\@ We consider it crucial to distinguish between valid IPA, for which it
is sufficient that any phonetically-trained reader is able to understand the
transcription, and strict IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are
aware of the following non-unique encoding options in the IPA, which will be
discussed in turn below:

@@

Most importantly though, in comparison to normal Unicode processing, the tie-bar
actually takes a reversed approach to complex graphemes. Basically, the Unicode
principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
sequences in a writing system have to be specified as tailored grapheme
clusters. Only in case the sequence is not a cluster, then this has to be
explicitly indicated. IPA takes a different approach. In IPA by default
different base letters are not connected into larger clusters; only when it is
specified in the string itself (using the tie bar).

@@

A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
the dark <l> in English codas, using the character <ɫ> \textsc{latin small
letter l with middle tilde} at \uni{026B}. This character is not canonically
equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
encodings has to be chosen. Because the tilde overlay is described as a general
mechanism by the IPA, we propose to use the separated <\dia{0334}>
\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
and avoid the tilde overlay completely.

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

@@

\noindent At the end of this chapter we have added a few longish tables summarizing all
159 different unicode codepoints that form the basis of strict-IPA encoding
(107 letters, 36 diacritics and 16 remaining symbols). Each of
these tables shows a typical glyph, and then lists the Unicode Codepoint,
Unicode Name and IPA description for each symbol. Further, there is a table with 
the additional options for valid-IPA and a table with the additional options for 
widened-IPA.\@

>>>>

As long as a transcription is only directed towards phonetically trained eyes,
then all the details of the Unicode-encoding are unimportant. For a linguist
reading an IPA transcription, many of the details that will be discussed in this
chapter might seem like hair-splitting trivialities. However, if IPA
transcriptions are intended to be used across resources (e.g.~searching similar
phenomena across different languages) then it becomes crucial that there are
strict encoding guidelines. Our main goal in this chapter is to present the
encoding issues and propose recommendations for a ``strict'' IPA encoding for
situations in which cross-resource consistency is crucial.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}. The 159 characters from 12 
code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
as discussed in Section~\ref{ipa-recommendations}.}

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the ``real'' retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        bevavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be used, respectively.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is the character to by used for a major group 
           break (by intonation), to be encoded by \textsc{double vertical line} 
           at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable when the goal is interoperability between resources written in
IPA.\@ We consider it crucial to distinguish between ``valid'' IPA, for which it
is sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are
aware of the following non-unique encoding options in the IPA, which will be
discussed in turn below:

@@

Most importantly though, in comparison to normal Unicode processing, the tie-bar
actually takes a reversed approach to complex graphemes. Basically, the Unicode
principle (see Section~\ref{pitfall-characters-are-not-graphemes}) is that fixed
sequences in a writing system have to be specified as ``tailored'' grapheme
clusters. Only in case the sequence is not a cluster, then this has to be
explicitly indicated. IPA takes a different approach. In IPA by default
different base letters are not connected into larger clusters; only when it is
specified in the string itself (using the tie bar).

@@

A similar situation arises with <\dia{0334}> \textsc{combining tilde overlay} at
\uni{0334}. The main reason some phoneticians like to use this in IPA is to mark
the ``dark'' <l> in English codas, using the character <ɫ> \textsc{latin small
letter l with middle tilde} at \uni{026B}. This character is not canonically
equivalent to the combination <l>~+~<\dia{0334}>, so one of the two possible
encodings has to be chosen. Because the tilde overlay is described as a general
mechanism by the IPA, we propose to use the separated <\dia{0334}>
\textsc{combining tilde overlay} at \uni{0334}. However, note that phonetically 
this seems to be (almost) superfluous (see Section~\ref{pitfall-multiple-options-ipa}) 
and the typical usage in the form of <ɫ> is (almost) a homoglyph with <ɬ> (see 
Section~\ref{pitfall-homoglyphs-in-IPA}). For these reasons we also suggest to try 
and avoid the tilde overlay completely.

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive ``sucking'' sound similar
      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ``ejective'' or ``creaky voicing'' are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

@@

\noindent At the end of this chapter we have added a few longish tables summarizing all
159 different unicode codepoints that form the basis of strict-IPA encoding
(107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
Each of these tables shows a typical glyph, and then lists the Unicode Codepoint,
Unicode Name and IPA description for each symbol. Further, there is a table with 
the additional options for valid-IPA and a table with the additional options for 
widened-IPA.\@
